---
title: "River Remedy Visualizations"
format: html
echo: true
toc: false
theme: lux
---

# Background

This project visualizes the results of the Fundación I.T.A. 2006 study of the Pilcomayo River basin in Chuquisaca, Bolivia. The focus is on concentrations of heavy metals in water, soil, sediment, vegetation, fish, human, and animal samples. This forms a base for more advanced exploration, spatial mapping, and statistical analysis in your summer research internship.

# Setup

```{r,include=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Water Samples

```{r,include=FALSE}
water <- read_csv("data/ITA_water_2006.csv")
glimpse(water)
```

## Lead in Water

```{r}
ggplot(water, aes(x = Location, y = `Pb (mg/l)`)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Lead Concentration in Water Samples",
    x = "Location",
    y = "mg/L"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Mercury in Water

```{r}
ggplot(water, aes(x = Location, y = `Hg (mg/l)`)) +
  geom_col(fill = "orange") +
  labs(
    title = "Mercury Concentration in Water Samples",
    x = "Location",
    y = "mg/L"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Soil Samples

```{r,include=FALSE}
soil <- read_csv("data/ITA_soil_2006.csv")
glimpse(soil)
```

## Zinc in Soil

```{r}
ggplot(soil, aes(x = Location, y = `Zn (mg/kg)`)) +
  geom_col(fill = "darkgreen") +
  labs(
    title = "Zinc Concentration in Soil Samples",
    x = "Location",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Arsenic in Soil

```{r}
ggplot(soil, aes(x = Location, y = `As (mg/kg)`)) +
  geom_col(fill = "red") +
  labs(
    title = "Arsenic Concentration in Soil Samples",
    x = "Location",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Sediment Samples

```{r,include=FALSE}
sediment <- read_csv("data/ITA_sed_2006.csv")
glimpse(sediment)
```

## Zinc in Sediments

```{r}
ggplot(sediment, aes(x = Location, y = `Zn (mg/kg)`)) +
  geom_col(fill = "purple") +
  labs(
    title = "Zinc Concentration in Sediment Samples",
    x = "Location",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Vegetation Samples

```{r,include=FALSE}
veg <- read_csv("data/ITA_veg_2006.csv")
glimpse(veg)
```

## Lead in Crops

```{r}
ggplot(veg, aes(x = Crop, y = `Pb (mg/kg)`)) +
  geom_col(fill = "brown") +
  labs(
    title = "Lead Concentration in Vegetables",
    x = "Crop Type",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Fish Samples

```{r,include=FALSE}
fish <- read_csv("data/ITA_fish_2006.csv")
glimpse(fish)
```

## Lead by Fish Tissue

```{r}
ggplot(fish, aes(x = `Sample Type`, y = `Pb (mg/kg)`)) +
  geom_col(fill = "red") +
  labs(
    title = "Lead in Fish Tissues",
    x = "Tissue Type",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Human Blood Samples

```{r,include=FALSE}
human <- read_csv("data/ITA_human_2006.csv")
glimpse(human)
```

## Lead in Children

```{r}
ggplot(human, aes(x = Location, y = `Pb [µg/dl] Children`)) +
  geom_col(fill = "pink") +
  labs(
    title = "Lead Concentration in Human Blood (Children)",
    x = "Location",
    y = "µg/dl"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Lead in Adults

```{r}
ggplot(human, aes(x = Location, y = `Pb [µg/dl] Adults`)) +
  geom_col(fill = "darkred") +
  labs(
    title = "Lead Concentration in Human Blood (Adults)",
    x = "Location",
    y = "µg/dl"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Animal Blood Samples

```{r,include=FALSE}
animal <- read_csv("data/ITA_animal_2006.csv")
glimpse(animal)
```

## Lead in Animals

```{r}
ggplot(animal, aes(x = Animal, y = `Pb (mg/dL)`)) +
  geom_col(fill = "darkblue") +
  labs(
    title = "Lead Concentration in Animal Blood Samples",
    x = "Animal Type",
    y = "mg/dL"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```
