---
title: "Bolivia Pilcomayo Heavy Metals Study"
author: "River Remedy - Katerina Bischel"
date: "2025-07-02"
format: html
toc: true
toc-depth: 3
number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

The following analysis explores heavy metal concentrations measured in 2006 across water, soil, sediment, vegetation, fish, humans, and animals in the Pilcomayo River basin of Chuquisaca, Bolivia. Data is derived from Fundación I.T.A. and serves as a baseline for environmental and health risk evaluation in your internship research.

# Objectives

-   Summarize heavy metal concentrations across environmental media
-   Visualize spatial and temporal variation
-   Identify sites exceeding threshold levels
-   Provide initial evidence for policy recommendations

# Methods

```{r}
library(tidyverse)
library(readr)
library(sf)
library(ggthemes)
library(patchwork)
```

Data were read from cleaned CSV files. Exploratory data analysis included descriptive statistics, barplots, and potential outlier detection.

# Data Summaries

## Water

```{r}
water <- read_csv("data/ITA_water_2006.csv")
```

```{r}
summary(water)
```

### Lead in Water

```{r}
ggplot(water, aes(x = reorder(Location, `Pb (mg/l)`), y = `Pb (mg/l)`)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Lead Concentration by Location (Water)", x = "Location", y = "mg/l") +
  theme_minimal()
```

### Mercury vs Lead

```{r}
ggplot(water, aes(x = `Pb (mg/l)`, y = `Hg (mg/l)`, label = Location)) +
  geom_point(size=3, color="darkred") +
  geom_text(hjust=1.2, vjust=0.5) +
  labs(title="Relationship of Lead and Mercury in Water", x="Pb (mg/l)", y="Hg (mg/l)") +
  theme_minimal()
```

## Soil

```{r}
soil <- read_csv("data/ITA_soil_2006.csv")
```

### Zinc Distribution in Soil

```{r}
ggplot(soil, aes(x = `Zn (mg/kg)`)) +
  geom_histogram(bins=15, fill="darkgreen", color="white") +
  labs(title = "Zinc Distribution in Soil Samples", x = "Zn (mg/kg)", y = "Count") +
  theme_minimal()

```

### Lead in Soil by Location

```{r}
ggplot(soil, aes(x = Location, y = `Pb (mg/kg)`)) +
  geom_boxplot(fill="brown") +
  labs(title = "Lead Concentrations in Soil", x = "Location", y = "mg/kg") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Sediment

```{r}
sediment <- read_csv("data/ITA_sed_2006.csv")
```

### Sediment Metals Summary

```{r}
sediment %>%
  select(Location, Zn = `Zn (mg/kg)`, Pb = `Pb (mg/kg)`, As = `As (mg/kg)`) %>%
  pivot_longer(-Location) %>%
  ggplot(aes(x = Location, y = value, fill = name)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="Sediment Heavy Metals by Location", y="mg/kg") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Vegetation

```{r}
veg <- read_csv("data/ITA_veg_2006.csv")
```

### Vegetation Metals by Crop Type

```{r}
veg %>%
  pivot_longer(cols = starts_with("Pb"), names_to="metal", values_to="value") %>%
  ggplot(aes(x=Crop, y=value, fill=metal)) +
  geom_col(position="dodge") +
  labs(title="Vegetation Lead Concentration by Crop", y="mg/kg") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Fish

```{r}
fish <- read_csv("data/ITA_fish_2006.csv")
```

### Lead in Fish Tissues

```{r}
ggplot(fish, aes(x = `Sample Type`, y = `Pb (mg/kg)`)) +
  geom_col(fill="dodgerblue") +
  labs(title="Lead Concentration in Fish Tissues", x="Tissue", y="mg/kg") +
  theme_minimal()
```

## Human

```{r}
human <- read_csv("data/ITA_human_2006.csv")
```

### Blood Lead Levels in Children

```{r}
ggplot(human, aes(x=Location, y=`Pb [µg/dl] Children`)) +
  geom_boxplot(fill="pink") +
  labs(title="Blood Lead Levels in Children", y="µg/dl") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Animals

```{r}
animal <- read_csv("data/ITA_animal_2006.csv")
```

### Animal Blood Lead Levels

```{r}
ggplot(animal, aes(x=Animal, y=`Pb (mg/dL)`)) +
  geom_col(fill="darkblue") +
  labs(title="Animal Blood Lead Concentrations", y="mg/dL") +
  theme_minimal()
```

# Exploratory Correlations
