---
title: "Bolivia Pilcomayo Heavy Metals Study - with WHO standards"
author: "River Remedy - Katerina Bischel"
date: "2025-07-02"
format: html
toc: true
toc-depth: 3
number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This analysis explores heavy metal concentrations from 2006 across the Pilcomayo River basin in Chuquisaca, Bolivia. The work is inspired by data from Fundación I.T.A. and aligned with WHO drinking-water guidelines and agricultural soil reference thresholds, with the aim of highlighting exceedances.

**Citations:**

-   World Health Organization. Guidelines for drinking-water quality, fourth edition incorporating the first addendum. Geneva: WHO, 2017. ISBN 978-92-4-154995-0. <https://www.who.int/publications/i/item/9789241549950>\
-   FAO/WHO Codex Alimentarius Standard 193-1995 Rev. 2018. General standard for contaminants and toxins in food and feed. <https://www.fao.org/fao-who-codexalimentarius>

# Setup

```{r}
library(tidyverse)
library(readr)
library(sf)
```

# Water Samples

```{r}
water <- read_csv("data/ITA_water_2006.csv")
```

## Lead in Water

```{r}
ggplot(water, aes(x = Location, y = `Pb (mg/l)`)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0.01, color = "red", linetype = "dashed") +
  annotate("text", x=Inf, y=0.01, label="WHO Pb limit 0.01 mg/L", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title = "Lead Concentration in Water vs WHO Limit",
    x = "Location",
    y = "mg/L"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Mercury in Water

```{r, fig.height=6}
ggplot(water, aes(x = Location, y = `Hg (mg/l)`)) +
  geom_col(fill = "orange") +
  geom_hline(yintercept = 0.006, color = "red", linetype = "dashed") +
  annotate("text", x=Inf, y=0.006, label="WHO Hg limit 0.006 mg/L", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title = "Mercury Concentration in Water vs WHO Limit",
    x = "Location",
    y = "mg/L"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Arsenic in Water

```{r}
ggplot(water, aes(x = Location, y = `As (mg/l)`)) +
  geom_col(fill = "purple") +
  geom_hline(yintercept = 0.01, color = "red", linetype = "dashed") +
  annotate("text", x=Inf, y=0.01, label="WHO As limit 0.01 mg/L", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title = "Arsenic Concentration in Water vs WHO Limit",
    x = "Location",
    y = "mg/L"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Cadmium in Water

```{r}
ggplot(water, aes(x = Location, y = `Cd (mg/l)`)) +
  geom_col(fill = "darkgreen") +
  geom_hline(yintercept = 0.003, color = "red", linetype = "dashed") +
  annotate("text", x=Inf, y=0.003, label="WHO Cd limit 0.003 mg/L", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title = "Cadmium Concentration in Water vs WHO Limit",
    x = "Location",
    y = "mg/L"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Soil Samples

```{r}
soil <- read_csv("data/ITA_soil_2006.csv")
```

## Lead in Soil

```{r}
ggplot(soil, aes(x = Location, y = `Pb (mg/kg)`)) +
  geom_col(fill = "brown") +
  geom_hline(yintercept = 70, color = "red", linetype = "dashed") +
  annotate("text", x=Inf, y=70, label="Soil Pb limit 70 mg/kg", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title = "Lead Concentration in Soil vs Reference Limit",
    x = "Location",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Zinc in Soil

```{r, fig.height=6}
ggplot(soil, aes(x = Location, y = `Zn (mg/kg)`)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 300, color = "red", linetype = "dashed") +
  annotate("text", x=Inf, y=300, label="Soil Zn ref 300 mg/kg", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title = "Zinc Concentration in Soil vs Reference",
    x = "Location",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Sediment: Lead

```{r}
sediment <- read_csv("data/ITA_sed_2006.csv")

```

```{r, fig.height=6}
ggplot(sediment, aes(x=Location, y=`Pb (mg/kg)`)) +
  geom_col(fill="orange") +
  geom_hline(yintercept=70, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=70, label="Ref Pb 70 mg/kg", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Lead in Sediment vs Reference",
    y="mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Sediment: Arsenic

```{r}
ggplot(sediment, aes(x=Location, y=`As (mg/kg)`)) +
  geom_col(fill="purple") +
  geom_hline(yintercept=20, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=20, label="Ref As 20 mg/kg", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Arsenic in Sediment vs Reference",
    y="mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Vegetation: Lead

```{r}
veg <- read_csv("data/ITA_veg_2006.csv")

```

```{r}
ggplot(veg, aes(x=Crop, y=`Pb (mg/kg)`)) +
  geom_col(fill="darkgreen") +
  geom_hline(yintercept=0.3, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=0.3, label="Codex Pb 0.3 mg/kg", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Lead in Crops vs Codex Limit",
    y="mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Vegetation: Cadmium

```{r}
ggplot(veg, aes(x=Crop, y=`Cd (mg/kg)`)) +
  geom_col(fill="brown") +
  geom_hline(yintercept=0.1, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=0.1, label="Codex Cd 0.1 mg/kg", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Cadmium in Crops vs Codex Limit",
    y="mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Fish: Lead

```{r}
fish <- read_csv("data/ITA_fish_2006.csv")

```

```{r}
ggplot(fish, aes(x=`Sample Type`, y=`Pb (mg/kg)`)) +
  geom_col(fill="dodgerblue") +
  geom_hline(yintercept=0.3, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=0.3, label="Codex Pb 0.3 mg/kg", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Lead in Fish vs Codex Limit",
    y="mg/kg"
  ) +
  theme_minimal()
```

## Human: Lead in Children

```{r}
human <- read_csv("data/ITA_human_2006.csv")

```

```{r}
ggplot(human, aes(x=Location, y=`Pb [µg/dl] Children`)) +
  geom_col(fill="pink") +
  geom_hline(yintercept=5, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=5, label="CDC ref 5 µg/dL", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Lead in Children Blood vs CDC Ref",
    y="µg/dL"
  ) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Human: Lead in Adults

```{r}
ggplot(human, aes(x=Location, y=`Pb [µg/dl] Adults`)) +
  geom_col(fill="darkred") +
  geom_hline(yintercept=10, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=10, label="CDC ref 10 µg/dL", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Lead in Adults Blood vs CDC Ref",
    y="µg/dL"
  ) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Animals: Lead

```{r}
animal <- read_csv("data/ITA_animal_2006.csv")

```

```{r}
ggplot(animal, aes(x=Animal, y=`Pb (mg/dL)`)) +
  geom_col(fill="darkblue") +
  geom_hline(yintercept=0.3, color="red", linetype="dashed") +
  annotate("text", x=Inf, y=0.3, label="Vet concern ~0.3 mg/dL", hjust=1.1, vjust=-0.5, color="red", size=3) +
  labs(
    title="Lead in Animal Blood vs Vet Concern Level",
    y="mg/dL"
  ) +
  theme_minimal()
```
