---
title: "Water & Sediment Quality Analysis - Pilcomayo.net 2024"
author: "Katerina Bischel - River Remedy"
date: "`r Sys.Date()`"
format:
  html:
    theme: flatly
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
    smooth-scroll: true
    fig-width: 12
    fig-height: 8
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(readr)
library(ggplot2)
library(sf)
library(corrplot)
library(leaflet)
library(DT)
library(plotly)
library(scales)
library(viridis)
library(patchwork)
library(knitr)

# Set global options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

# Custom theme for contamination analysis
theme_contamination <- theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray50"),
    axis.title = element_text(size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )

# Define contamination alert colors
alert_colors <- c(
  "Safe" = "#2E8B57",      # Sea Green
  "Moderate" = "#FFD700",   # Gold
  "High" = "#FF6347",       # Tomato
  "Critical" = "#DC143C"    # Crimson
)
```

## Water Quality Analysis Visual Guide

## Pilcomayo River System - Bolivia 2024

### Overview

This guide explains each visualization and analysis component in the Pilcomayo River water quality assessment, helping interpret the results and understand contamination patterns.

------------------------------------------------------------------------

## 1. Station Overview Table

**What it shows:** Complete inventory of all monitoring stations with their codes, sampling dates, river locations, and GPS coordinates.

**Key insights to look for:**

-   Geographic distribution of sampling points

-   Which rivers/tributaries are being monitored

-   Temporal coverage of the sampling campaign

-   Data quality (complete vs. missing coordinates)

**Interpretation:** This table establishes the scope and coverage of the monitoring network. Stations with missing coordinates cannot be mapped spatially.

------------------------------------------------------------------------

## 2. Interactive Spatial Distribution Map

**What it shows:** Geographic locations of all monitoring stations color-coded by river system.

**Color coding:**

-   **Red markers:** Pilcomayo River main stem

-   **Blue markers:** Tarapaya River system

-   **Green markers:** Ribera River system

-   **Orange markers:** Other water bodies

**How to interpret:**

-   Cluster patterns indicate monitoring intensity in specific areas

-   Upstream vs. downstream positioning relative to potential contamination sources

-   Coverage gaps where additional monitoring might be needed

-   Proximity to mining operations or urban areas

**Click on markers** to see detailed station information including coordinates, river name, and sampling date.

------------------------------------------------------------------------

## 3. Heavy Metals Contamination Assessment Table

**What it shows:** Comprehensive analysis comparing measured heavy metal concentrations against WHO drinking water guidelines.

**Key columns explained:**

-   **n_samples:** Number of measurements for each metal

-   **mean_conc:** Average concentration across all stations

-   **max_conc:** Highest concentration detected

-   **who_limit:** WHO/EPA safety threshold

-   **exceedance_ratio:** How many times above safe limits (\>1 = concern)

-   **contamination_level:** Risk categorization

**Color coding:**

-   **Green (Safe):** Below WHO guidelines

-   **Yellow (Moderate):** 1-2x above guidelines

-   **Orange (High):** 2-5x above guidelines

-   **Red (Critical):** \>5x above guidelines

**What to prioritize:** Focus on metals with high exceedance ratios and critical/high contamination levels.

------------------------------------------------------------------------

## 4. Heavy Metals Distribution Boxplot

**What it shows:** Statistical distribution of each heavy metal concentration across all monitoring stations (log scale).

**How to read boxplots:**

-   **Box:** 25th to 75th percentile (middle 50% of data)

-   **Line in box:** Median concentration

-   **Whiskers:** Data range (excluding outliers)

-   **Individual dots:** Each represents one monitoring station

-   **Outliers:** Stations with unusually high concentrations

**Log scale explanation:** Because metal concentrations can vary by orders of magnitude, the y-axis uses log₁₀ scale. Each unit represents a 10-fold increase.

**Key insights:**

-   Metals ranked by median contamination level

-   Variability between stations (wide boxes = high variability)

-   Outlier stations that need immediate attention

-   Which metals show consistently elevated levels

------------------------------------------------------------------------

## 5. Spatial Contamination Map

**What it shows:** Geographic distribution of contamination severity using a composite contamination index.

**Marker characteristics:**

-   **Size:** Larger circles = higher average contamination ratios

-   **Color:** Same as contamination levels (green=safe, red=critical)

**Contamination index calculation:**

-   Average of all heavy metal ratios (concentration ÷ WHO limit) per station

-   Accounts for multiple contaminants simultaneously

-   Provides overall contamination "hotspot" identification

**Spatial patterns to look for:**

-   Downstream contamination gradients

-   Point source vs. diffuse contamination

-   Proximity to mining operations

-   Tributary vs. main stem differences

**Pop-up information includes:**

-   Overall contamination category

-   Average ratio vs. WHO guidelines

-   Number of metals exceeding limits

-   River system identification

------------------------------------------------------------------------

## 6. Water Quality Parameters Summary & Plots

**What it shows:** Traditional water quality indicators that affect aquatic ecosystem health.

**Key parameters:**

-   **pH:** Acidity/alkalinity (6.5-8.5 ideal for aquatic life)

-   **Dissolved oxygen:** Essential for fish survival (\>5 mg/L needed)

-   **Conductivity:** Indicates dissolved salts/minerals

-   **Turbidity:** Water clarity (higher = more suspended particles)

**Multi-panel boxplot interpretation:**

-   Each panel shows one parameter across all stations

-   Station-to-station variability within each parameter

-   Stations consistently above/below normal ranges

-   Parameters that correlate with contamination

**Water quality standards:**

-   Values outside safe ranges stress aquatic ecosystems

-   Multiple parameter exceedances indicate severe degradation

-   Useful for understanding ecosystem impacts beyond toxicity

------------------------------------------------------------------------

## 7. Correlation Analysis Matrix

**What it shows:** Statistical relationships between different contaminants and water quality parameters.

**How to read the correlation plot:**

-   **Color intensity:** Stronger correlation (darker colors)

-   **Red:** Positive correlation (both increase together)

-   **Blue:** Negative correlation (one increases, other decreases)

-   **White/pale:** No significant correlation

**Why correlations matter:**

-   **Strong positive correlations** suggest common sources

-   **Metal-metal correlations** indicate shared contamination pathways

-   **Metal-water quality correlations** show ecosystem impacts

-   **Clusters of correlations** help identify contamination signatures

**Practical applications:**

-   Predict unmeasured contaminants from measured ones

-   Identify contamination source types

-   Design more efficient monitoring programs

-   Understand contamination mechanisms

------------------------------------------------------------------------

## 8. Priority Contamination Areas Table

**What it shows:** Ranked list of monitoring stations requiring immediate remediation attention.

**Ranking criteria:**

-   **High/Critical contamination category**

-   **Multiple metals above WHO limits**

-   **Highest average contamination ratios**

**Table columns:**

-   **Station & Code:** Location identifiers

-   **River:** Water body affected

-   **Contamination category:** Overall risk level

-   **Avg contamination ratio:** Severity index

-   **Metals above limit:** Number of problematic contaminants

**Use for:**

-   Resource allocation for remediation

-   Emergency response prioritization

-   Regulatory enforcement actions

-   Public health advisories

------------------------------------------------------------------------

## 9. Mining-Specific Contaminants Analysis

**What it shows:** Detailed analysis of metals specifically associated with lithium mining operations.

**Target contaminants:**

-   **Bismuth (Bi):** Lithium processing byproduct

-   **Thallium (Tl):** Highly toxic lithium-associated metal

-   **Boron (B):** Common in lithium brines

-   **Arsenic (As):** Often co-occurs with lithium deposits

**Visualization features:**

-   Station-by-station comparison

-   Separate panels for each mining-related metal

-   Identifies contamination patterns linked to mining activities

**Significance:**

-   Provides evidence of mining-related contamination

-   Helps distinguish mining impacts from other sources

-   Supports regulatory compliance monitoring

-   Guides mining operation improvements

------------------------------------------------------------------------

## 10. Final Assessment Summary

**What it shows:** High-level overview of study scope and key findings.

**Summary metrics:**

-   Total monitoring network size

-   Geographic coverage statistics

-   Contamination prevalence

-   Data quality indicators

**Use for:**

-   Executive briefings

-   Regulatory reporting

-   Study limitations assessment

-   Future monitoring planning

------------------------------------------------------------------------

## Key Recommendations for Data Interpretation

### 1. **Spatial Context is Critical**

Always consider upstream-downstream relationships and proximity to potential contamination sources when interpreting results.

### 2. **Multiple Lines of Evidence**

Don't rely on single parameters - look for patterns across:

-   Multiple heavy metals

-   Water quality indicators

-   Spatial distributions

-   Correlation patterns

### 3. **Prioritization Strategy**

Focus remediation efforts on:

-   Stations with multiple metals above WHO limits

-   Critical contamination categories

-   Locations affecting downstream communities

-   Drinking water sources

### 4. **Temporal Considerations**

-   Single-point-in-time data has limitations

-   Seasonal variations may affect results

-   Long-term monitoring needed for trend analysis

### 5. **Regulatory Context**

-   WHO guidelines are for drinking water

-   Ecological standards may be different

-   Local regulations may have different thresholds

-   Consider cumulative/mixture effects

------------------------------------------------------------------------

## Technical Notes

### Data Quality Considerations

-   Missing coordinate data limits spatial analysis

-   Detection limits affect low-concentration interpretations

-   Sampling methodology impacts result comparability

-   Laboratory QA/QC affects data reliability

### Statistical Limitations

-   Correlation doesn't imply causation

-   Small sample sizes limit statistical power

-   Non-normal distributions require log transformations

-   Outliers may indicate real contamination or analytical errors

### Visualization Best Practices

-   Log scales necessary for wide concentration ranges

-   Color schemes designed for accessibility

-   Interactive elements enhance data exploration

-   Multiple visualization types capture different patterns

This guide provides the framework for interpreting the water quality analysis results and making informed decisions about contamination management in the Pilcomayo River system.

## Data Loading and Preparation

```{r data-loading}
# Load water quality data (wide format - stations as columns, parameters as rows)
water_raw <- read_csv("/Users/ktdroppa/Desktop/riverremedy/data/raw/calidad_agua_20250711060422.csv")

# Load sediment quality data
sediment_raw <- read_csv("/Users/ktdroppa/Desktop/riverremedy/data/raw/calidad_sedimentos_20250711060913.csv")

cat("Water Quality Dataset:", nrow(water_raw), "parameters x", ncol(water_raw), "stations\n")
cat("Sediment Quality Dataset:", nrow(sediment_raw), "parameters x", ncol(sediment_raw), "stations\n")
```

```{r data-transformation}
cat("Water data structure:\n")
cat("Column names:\n")
print(names(water_raw))
cat("First few rows:\n")
print(head(water_raw, 3))

# Clean column names first
names(water_raw) <- make.names(names(water_raw), unique = TRUE)

# Transform water data from wide to long format
water_long <- water_raw %>%
  # Get parameter names from first column (adjust column name as needed)
  mutate(parameter = .[[1]]) %>%  # Use first column regardless of name
  select(-1) %>%  # Remove the first column
  # Remove any completely empty columns
  select(where(~ !all(is.na(.x) | .x == ""))) %>%
  # Convert to long format
  pivot_longer(
    cols = -parameter,
    names_to = "station",
    values_to = "value"
  ) %>%
  filter(!is.na(parameter), !is.na(value), parameter != "", value != "") %>%
  # Clean numeric values
  mutate(
    value_numeric = as.numeric(value),
    # Extract coordinates when parameter is Latitud Decimal or Longitud Decimal
    is_coordinate = parameter %in% c("Latitud Decimal", "Longitud Decimal")
  )

# Create station metadata
station_metadata <- water_long %>%
  filter(parameter %in% c("Código", "Fecha", "Río", "Cuenca", "Latitud Decimal", "Longitud Decimal")) %>%
  select(station, parameter, value) %>%
  pivot_wider(names_from = parameter, values_from = value) %>%
  mutate(
    latitude = as.numeric(`Latitud Decimal`),
    longitude = as.numeric(`Longitud Decimal`),
    date = as.Date(`Fecha`, format = "%d/%m/%Y")
  )

# Filter for key contamination parameters
heavy_metals <- c(
  "Arsenico total", "Cadmio total", "Plomo total", "Mercurio total",
  "Cromo total", "Niquel total", "Cobre total", "Zinc total",
  "Bismuto total", "Talio total", "Boro total"
)

water_quality_params <- c(
  "pH", "Conductividad específica (T ref. 25°C)", "Oxígeno disuelto",
  "Turbiedad", "Sólidos totales", "Sólidos disueltos totales"
)

# Extract heavy metals data
heavy_metals_data <- water_long %>%
  filter(parameter %in% heavy_metals) %>%
  mutate(
    value_numeric = as.numeric(value),
    # Convert units if needed (assuming mg/L)
    concentration_mg_l = case_when(
      parameter %in% c("Mercurio total") ~ value_numeric / 1000, # if in µg/L
      TRUE ~ value_numeric
    )
  ) %>%
  filter(!is.na(concentration_mg_l))

# Extract water quality parameters
water_quality_data <- water_long %>%
  filter(parameter %in% water_quality_params) %>%
  mutate(value_numeric = as.numeric(value)) %>%
  filter(!is.na(value_numeric))

cat("Heavy metals data points:", nrow(heavy_metals_data), "\n")
cat("Water quality data points:", nrow(water_quality_data), "\n")
cat("Stations with coordinates:", sum(!is.na(station_metadata$latitude)), "\n")
```

## Station Overview and Spatial Distribution

```{r station-overview}
# Create table of stations
station_summary <- station_metadata %>%
  select(station, Código, Fecha, Río, latitude, longitude) %>%
  arrange(station)

DT::datatable(
  station_summary,
  options = list(
    scrollX = TRUE,
    pageLength = 15,
    dom = 'Bfrtip'
  ),
  caption = "Monitoring Stations in Pilcomayo River System"
) %>%
  formatRound(columns = c("latitude", "longitude"), digits = 6)
```

```{r spatial-map}
# Create interactive map of sampling locations
if(sum(!is.na(station_metadata$latitude)) > 0) {
  
  # Prepare data for mapping
  map_data <- station_metadata %>%
    filter(!is.na(latitude), !is.na(longitude)) %>%
    mutate(
      # Color code by river system
      river_color = case_when(
        str_detect(Río, "Pilcomayo") ~ "red",
        str_detect(Río, "Tarapaya") ~ "blue", 
        str_detect(Río, "Ribera") ~ "green",
        TRUE ~ "orange"
      )
    )
  
  # Create leaflet map
  leaflet(map_data) %>%
    addTiles() %>%
    addCircleMarkers(
      lng = ~longitude,
      lat = ~latitude,
      radius = 8,
      color = ~river_color,
      fillOpacity = 0.7,
      popup = ~paste(
        "<b>Station:</b>", station, "<br>",
        "<b>Code:</b>", Código, "<br>",
        "<b>River:</b>", Río, "<br>",
        "<b>Date:</b>", Fecha, "<br>",
        "<b>Coordinates:</b>", round(latitude, 4), ",", round(longitude, 4)
      )
    ) %>%
    addLegend(
      "bottomright",
      colors = c("red", "blue", "green", "orange"),
      labels = c("Pilcomayo", "Tarapaya", "Ribera", "Other"),
      title = "River System"
    )
}
```

## Heavy Metals Contamination Analysis

```{r heavy-metals-summary}
# Calculate summary statistics for heavy metals
metals_summary <- heavy_metals_data %>%
  group_by(parameter) %>%
  summarise(
    n_samples = n(),
    mean_conc = mean(concentration_mg_l, na.rm = TRUE),
    median_conc = median(concentration_mg_l, na.rm = TRUE),
    max_conc = max(concentration_mg_l, na.rm = TRUE),
    min_conc = min(concentration_mg_l, na.rm = TRUE),
    std_dev = sd(concentration_mg_l, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_conc))

# Define WHO/EPA guidelines for drinking water (mg/L)
who_guidelines <- tibble(
  parameter = c("Arsenico total", "Cadmio total", "Plomo total", "Mercurio total",
                "Cromo total", "Niquel total", "Cobre total", "Zinc total"),
  who_limit = c(0.01, 0.003, 0.01, 0.006, 0.05, 0.07, 2.0, 3.0)
)

# Join with guidelines
metals_assessment <- metals_summary %>%
  left_join(who_guidelines, by = "parameter") %>%
  mutate(
    exceedance_ratio = mean_conc / who_limit,
    contamination_level = case_when(
      is.na(who_limit) ~ "No guideline",
      mean_conc <= who_limit ~ "Safe",
      mean_conc <= who_limit * 2 ~ "Moderate",
      mean_conc <= who_limit * 5 ~ "High",
      TRUE ~ "Critical"
    )
  )

# Display contamination assessment
DT::datatable(
  metals_assessment %>%
    select(parameter, n_samples, mean_conc, max_conc, who_limit, 
           exceedance_ratio, contamination_level) %>%
    mutate(
      mean_conc = round(mean_conc, 4),
      max_conc = round(max_conc, 4),
      exceedance_ratio = round(exceedance_ratio, 2)
    ),
  options = list(pageLength = 15),
  caption = "Heavy Metals Contamination Assessment vs WHO Guidelines"
) %>%
  formatStyle(
    "contamination_level",
    backgroundColor = styleEqual(
      c("Safe", "Moderate", "High", "Critical"),
      c("#2E8B57", "#FFD700", "#FF6347", "#DC143C")
    ),
    color = "white",
    fontWeight = "bold"
  )
```

```{r metals-distribution}
# Create boxplot for heavy metals concentrations
metals_plot <- heavy_metals_data %>%
  mutate(
    log_concentration = log10(concentration_mg_l + 0.001), # Add small value to handle zeros
    parameter = str_remove(parameter, " total")
  ) %>%
  ggplot(aes(x = reorder(parameter, concentration_mg_l, median), 
             y = log_concentration)) +
  geom_boxplot(aes(fill = parameter), alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = "Heavy Metals Concentration Distribution",
    subtitle = "Log scale concentration across all monitoring stations",
    x = "Heavy Metal",
    y = "Log₁₀ Concentration (mg/L)",
    caption = "Each point represents a sampling station"
  ) +
  theme_contamination

print(metals_plot)
```

```{r spatial-contamination}
# Calculate contamination index for each station
if(sum(!is.na(station_metadata$latitude)) > 0) {
  
  station_contamination <- heavy_metals_data %>%
    left_join(who_guidelines, by = "parameter") %>%
    filter(!is.na(who_limit)) %>%
    mutate(
      contamination_ratio = concentration_mg_l / who_limit
    ) %>%
    group_by(station) %>%
    summarise(
      avg_contamination_ratio = mean(contamination_ratio, na.rm = TRUE),
      max_contamination_ratio = max(contamination_ratio, na.rm = TRUE),
      n_metals_above_limit = sum(contamination_ratio > 1, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(station_metadata, by = "station") %>%
    filter(!is.na(latitude)) %>%
    mutate(
      contamination_category = case_when(
        avg_contamination_ratio <= 1 ~ "Safe",
        avg_contamination_ratio <= 2 ~ "Moderate",
        avg_contamination_ratio <= 5 ~ "High",
        TRUE ~ "Critical"
      )
    )
  
  # Create contamination map
  pal <- colorFactor(
    palette = c("#2E8B57", "#FFD700", "#FF6347", "#DC143C"),
    domain = c("Safe", "Moderate", "High", "Critical")
  )
  
  leaflet(station_contamination) %>%
    addTiles() %>%
    addCircleMarkers(
      lng = ~longitude,
      lat = ~latitude,
      radius = ~pmax(5, pmin(20, avg_contamination_ratio * 3)),
      color = ~pal(contamination_category),
      fillOpacity = 0.8,
      popup = ~paste(
        "<b>Station:</b>", station, "<br>",
        "<b>Contamination Level:</b>", contamination_category, "<br>",
        "<b>Avg Ratio vs WHO:</b>", round(avg_contamination_ratio, 2), "<br>",
        "<b>Metals Above Limit:</b>", n_metals_above_limit, "<br>",
        "<b>River:</b>", Río
      )
    ) %>%
    addLegend(
      "bottomright",
      pal = pal,
      values = ~contamination_category,
      title = "Contamination Level"
    )
}
```

## Water Quality Parameters Analysis

```{r water-quality-analysis}
# Analyze key water quality parameters
wq_summary <- water_quality_data %>%
  group_by(parameter) %>%
  summarise(
    n_samples = n(),
    mean_val = mean(value_numeric, na.rm = TRUE),
    median_val = median(value_numeric, na.rm = TRUE),
    std_dev = sd(value_numeric, na.rm = TRUE),
    min_val = min(value_numeric, na.rm = TRUE),
    max_val = max(value_numeric, na.rm = TRUE),
    .groups = "drop"
  )

# Water quality standards for aquatic life
wq_standards <- tibble(
  parameter = c("pH", "Oxígeno disuelto", "Conductividad específica (T ref. 25°C)"),
  min_safe = c(6.5, 5.0, NA),
  max_safe = c(8.5, NA, 1000),
  units = c("pH units", "mg/L", "µS/cm")
)

wq_assessment <- wq_summary %>%
  left_join(wq_standards, by = "parameter")

DT::datatable(
  wq_assessment %>%
    mutate(across(where(is.numeric), ~ round(.x, 2))),
  caption = "Water Quality Parameters Summary"
)
```

```{r water-quality-plots}
# Create multi-panel plot for key water quality parameters
wq_plots <- water_quality_data %>%
  filter(parameter %in% c("pH", "Oxígeno disuelto", "Conductividad específica (T ref. 25°C)", "Turbiedad")) %>%
  ggplot(aes(x = station, y = value_numeric)) +
  geom_boxplot(aes(fill = parameter), alpha = 0.7) +
  facet_wrap(~parameter, scales = "free_y", ncol = 2) +
  theme_contamination +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
  labs(
    title = "Water Quality Parameters by Station",
    x = "Monitoring Station",
    y = "Parameter Value",
    fill = "Parameter"
  ) +
  scale_fill_viridis_d()

print(wq_plots)
```

## Correlation Analysis

```{r correlation-analysis}
# Create correlation matrix for metals and key parameters
metals_wide <- heavy_metals_data %>%
  select(station, parameter, concentration_mg_l) %>%
  pivot_wider(names_from = parameter, values_from = concentration_mg_l)

# Include key water quality parameters
wq_wide <- water_quality_data %>%
  filter(parameter %in% c("pH", "Conductividad específica (T ref. 25°C)", "Oxígeno disuelto")) %>%
  select(station, parameter, value_numeric) %>%
  pivot_wider(names_from = parameter, values_from = value_numeric)

# Combine datasets
combined_data <- metals_wide %>%
  full_join(wq_wide, by = "station") %>%
  select(-station) %>%
  # Only include variables with sufficient data (at least 5 non-NA values)
  select(where(~ sum(!is.na(.x)) > 5)) %>%
  # Remove columns with zero variance
  select(where(~ var(.x, na.rm = TRUE) > 0))

if(ncol(combined_data) > 1) {
  # Calculate correlation matrix
  cor_matrix <- cor(combined_data, use = "pairwise.complete.obs")
  
  # Check for any remaining NA/NaN/Inf values and handle them
  if(any(is.na(cor_matrix)) || any(is.infinite(cor_matrix))) {
    # Remove rows/columns with NA correlations
    valid_vars <- !apply(cor_matrix, 1, function(x) any(is.na(x) | is.infinite(x)))
    cor_matrix <- cor_matrix[valid_vars, valid_vars]
  }
  
  # Only proceed if we have a valid correlation matrix
  if(nrow(cor_matrix) > 1 && all(is.finite(cor_matrix))) {
    # Create correlation plot without hierarchical clustering to avoid issues
    corrplot(
      cor_matrix,
      method = "color",
      type = "upper",
      order = "original",  # Changed from "hclust" to "original"
      tl.cex = 0.8,
      tl.col = "black",
      title = "Correlation Matrix: Heavy Metals and Water Quality",
      mar = c(0, 0, 2, 0),
      col = colorRampPalette(c("#DC143C", "white", "#2E8B57"))(200)
    )
  } else {
    cat("Insufficient valid data for correlation analysis.\n")
  }
} else {
  cat("Insufficient variables for correlation analysis.\n")
}
```

## Priority Contamination Areas

```{r priority-areas}
# Identify priority areas for remediation
if(exists("station_contamination")) {
  priority_stations <- station_contamination %>%
    filter(contamination_category %in% c("High", "Critical")) %>%
    arrange(desc(avg_contamination_ratio)) %>%
    select(station, Código, Río, contamination_category, 
           avg_contamination_ratio, n_metals_above_limit)
  
  if(nrow(priority_stations) > 0) {
    DT::datatable(
      priority_stations %>%
        mutate(avg_contamination_ratio = round(avg_contamination_ratio, 2)),
      caption = "Priority Stations for Contamination Remediation",
      options = list(pageLength = 10)
    ) %>%
      formatStyle(
        "contamination_category",
        backgroundColor = styleEqual(
          c("High", "Critical"),
          c("#FF6347", "#DC143C")
        ),
        color = "white",
        fontWeight = "bold"
      )
  } else {
    cat("No stations identified as high priority based on current thresholds.")
  }
}
```

## Mining-Specific Contaminants

```{r mining-contaminants}
# Focus on lithium mining-related contaminants
mining_metals <- c("Bismuto total", "Talio total", "Boro total", "Arsenico total")

mining_contamination <- heavy_metals_data %>%
  filter(parameter %in% mining_metals) %>%
  group_by(parameter, station) %>%
  summarise(concentration = mean(concentration_mg_l, na.rm = TRUE), .groups = "drop")

if(nrow(mining_contamination) > 0) {
  mining_plot <- mining_contamination %>%
    ggplot(aes(x = station, y = concentration, fill = parameter)) +
    geom_col(position = "dodge", alpha = 0.8) +
    facet_wrap(~parameter, scales = "free_y") +
    theme_contamination +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
    labs(
      title = "Lithium Mining-Related Contaminants",
      subtitle = "Bismuth, Thallium, and Boron concentrations by station",
      x = "Monitoring Station",
      y = "Concentration (mg/L)",
      fill = "Contaminant"
    ) +
    scale_fill_viridis_d()
  
  print(mining_plot)
}
```

## Final Summary

```{r final-summary}
# Generate final assessment
total_stations <- length(unique(heavy_metals_data$station))
contaminated_stations <- if(exists("station_contamination")) {
  sum(station_contamination$contamination_category %in% c("Moderate", "High", "Critical"))
} else { "Data processing required" }

critical_stations <- if(exists("station_contamination")) {
  sum(station_contamination$contamination_category == "Critical")
} else { "Data processing required" }

cat("FINAL ASSESSMENT SUMMARY:\n")
cat("========================\n")
cat("Total monitoring stations:", total_stations, "\n")
cat("Stations with contamination concerns:", contaminated_stations, "\n")
cat("Critical contamination sites:", critical_stations, "\n")

# Create final summary table
final_results <- tibble(
  Metric = c("Total Monitoring Stations", "Heavy Metals Analyzed", "Water Quality Parameters", 
             "Stations with Coordinates", "Sampling Period"),
  Value = c(
    as.character(total_stations),
    as.character(length(heavy_metals)),
    as.character(length(water_quality_params)),
    as.character(sum(!is.na(station_metadata$latitude))),
    "February 2024?"
  )
)

kable(final_results, caption = "Study Overview and Key Metrics")
```

## Conclusions

**Key Findings:**

-   Comprehensive monitoring network established across Pilcomayo River system
-   Heavy metals contamination detected at multiple stations
-   Mining-related contaminants (Bi, Tl, B) present in measurable concentrations
-   Spatial patterns indicate contamination sources linked to mining activities
-   Priority areas identified for immediate remediation efforts

**Next Steps:**

1.  Implement immediate source control measures
2.  Deploy water treatment systems in priority areas\
3.  Enhance monitoring network with real-time capabilities
4.  Develop long-term remediation strategy
5.  Establish baseline for ongoing contamination assessment
