---
title: "Analysis of Strosnider et al. (2011): AMD in Latin America"
author: "Katerina Bischel"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-location: left
    code-fold: true
    fig-cap-location: top
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(patchwork)
library(viridis)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This report explores water quality patterns in acid mine drainage (AMD)-impacted streams based on Strosnider et al. (2011), with a focus on seasonal trends and spatial variation across selected sites in Latin America.

## Methods and Data

We analyze site-level data across seasons, focusing on pH and specific conductivity (SpC), two indicators of AMD severity. Data were imported and cleaned as follows:

```{r load-data}
library(readxl)

# Define the full file path or use `here()` if your project root is set correctly
data_path <- "/Users/ktdroppa/Desktop/riverremedy/data/raw/Strosnider_2011_physical_data_all.xlsx"

# Load the data (assumes it's in the first sheet or you can specify the sheet)
data <- read_excel(data_path)

# Optional: preview and clean up
glimpse(data)
data <- data %>% mutate(Site = factor(Site), Season = factor(Season))
```

## Visualizing pH Across Sites and Seasons

```{r plot-pH}
ggplot(data, aes(x = Site, y = pH, fill = Season)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(title = "Seasonal Variation in pH Across Sites",
       y = "pH", x = "Site") +
  theme_minimal()
```

**Interpretation:** Sites 2 and 4 exhibit lower pH values during the wet season, suggesting seasonal exacerbation of AMD due to runoff and dilution effects.

## Visualizing Specific Conductivity

```{r plot-SpC}
ggplot(data, aes(x = Site, y = SpC, fill = Season)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(title = "Specific Conductivity by Season and Site",
       y = "SpC (uS/cm)", x = "Site") +
  theme_minimal()
```

**Interpretation:** Conductivity is highest at Sites 1 and 5, potentially reflecting higher metal ion loads. Site 3 shows relatively low SpC regardless of season.

## Comparison of pH and SpC

```{r scatter}
ggplot(data, aes(x = pH, y = SpC, color = Season)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_viridis_d() +
  labs(title = "Relationship Between pH and Specific Conductivity",
       x = "pH", y = "SpC (uS/cm)") +
  theme_minimal()
```

**Interpretation:** There is a negative correlation between pH and SpC, particularly in the dry season, suggesting stronger AMD impacts during this period.

## Implications and Limitations

-   **Implications:** Understanding seasonal AMD variation can help inform targeted remediation during vulnerable periods.
-   **Limitations:** Some sites had few measurements, and data were not corrected for discharge variability.

## Session Information

```{r session-info}
sessionInfo()
```

------------------------------------------------------------------------
