---
title: '01'
author: "Katerina Bischel"
date: "2025-07-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Background

This project visualizes results from the Fundación Instituto de Tecnología de Alimentos (I.T.A.) study conducted in 2006, focused on the Pilcomayo River basin in Chuquisaca, Bolivia. The study analyzed metals in water, soil, sediments, fish, human and animal blood, and plants.

Source: Muestreo de aguas, suelos, vegetales, sangre humana y animales, peces y sedimentos en puntos seleccionados en la cuenca del Río Pilcomayo en Chuquisaca (2006)

Libraries

```{r}
library(readr)
library(ggplot2)
library(dplyr)
```
Data Visualizations

Water Samples

```{r}
water <- read_csv("data/ITA_water_2006.csv")
head(water)
```
Lead in Water

```{r, fig.width = 15}
ggplot(water, aes(x = Location, y = `Pb (mg/l)`)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Lead Concentration in Water Samples",
    x = "Location",
    y = "mg/L"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Soil Samples

```{r}
soil <- read_csv("data/ITA_soil_2006.csv")
head(soil)
```

Zinc in Soil

```{r, fig.width= 10}
ggplot(soil, aes(x = Location, y = `Zn (mg/kg)`)) +
  geom_col(fill = "darkgreen") +
  labs(
    title = "Zinc Concentration in Soil Samples",
    x = "Location",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



Sediment Samples

```{r}
sediment <- read_csv("data/ITA_sed_2006.csv")
head(sediment)
```

Mercury in Sediments

```{r, fig.width=10}
ggplot(sediment, aes(x = Location, y = `Zn (mg/kg)`)) +
  geom_col(fill = "orange") +
  labs(
    title = "Zinc Concentration in Sediment Samples",
    x = "Location",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Vegetation Samples

```{r}
veg <- read_csv("data/ITA_veg_2006.csv")
head(veg)
```

Lead in Vegetables

```{r}
ggplot(veg, aes(x = Crop, y = `Zn (mg/kg)`)) +
  geom_col(fill = "purple") +
  labs(
    title = "Zinc Concentration in Vegetables",
    x = "Crop Type",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Fish Samples

```{r}
fish <- read_csv("data/ITA_fish_2006.csv")
head(fish)
```

Lead in Fish Muscle Tissue

```{r}
ggplot(fish, aes(x = `Sample Type`, y = `Pb (mg/kg)`)) +
  geom_col(fill = "red") +
  labs(
    title = "Lead Concentration in Fish Samples",
    x = "Sample Type",
    y = "mg/kg"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Human Blood Samples

```{r}
human <- read_csv("data/ITA_human_2006.csv")
head(human)
```

Lead in Child Human Blood

```{r}
ggplot(human, aes(x = Location, y = `Pb [µg/dl] Children`)) +
  geom_col(fill = "brown") +
  labs(
    title = "Lead Concentration in Human Blood (Children)",
    x = "Location",
    y = "µg/dl"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Lead in Adult Human Blood

```{r}
ggplot(human, aes(x = Location, y = `Pb [µg/dl] Adults`)) +
  geom_col(fill = "darkred") +
  labs(
    title = "Lead Concentration in Human Blood (Adults)",
    x = "Location",
    y = "µg/dl"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Animal Blood Samples

```{r}
animal <- read_csv("data/ITA_animal_2006.csv")
head(animal)
```

Lead in Animal Blood

```{r}
ggplot(animal, aes(x = Animal, y = `Pb (mg/dL)`)) +
  geom_col(fill = "darkred") +
  labs(
    title = "Lead Concentration in Animal Blood Samples",
    x = "Animal Type",
    y = "mg/dL"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Conclusions

This is a preliminary visualization of the 2006 Pilcomayo River metals assessment. Further statistical summaries and comparisons to WHO or Bolivian environmental standards can be added as needed.